---
title: "R Practice: Functions"
format: html
editor: visual
---

## Loading pacakges

```{r}
library(rvest)
library(readr)
library(dplyr)
library(janitor)
```

## Scraping species table from Bird Banding Lab

```{r}
webpage <- rvest::read_html("https://www.pwrc.usgs.gov/BBL/Bander_Portal/login/speclist.php")

tbls <- rvest::html_nodes(webpage, "table") %>% 
    rvest::html_table(fill = TRUE)

species <- tbls[[1]] %>% 
    janitor::clean_names() %>% 
    select(alpha_code, common_name) %>% 
    mutate(alpha_code = tolower(alpha_code))

head(species)
```

## Loading data from the [Arctic Data Center](https://arcticdata.io/catalog/view/doi:10.18739/A23R0PT35)

-   Utqiagvik_predator_surveys.csv
-   Utqiagvik_nest_data.csv
-   Utqiagvik_egg_measurements.csv

```{r}
#| message: false
nest_data <- read_csv("https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A982bd2fc-4edf-4da7-96ef-0d11b853102d")

predator_survey <- read_csv("https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A9ffec04c-7e2d-41dd-9e88-b6c2e8c4375e")

egg_measures <- read_csv("https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A4b219711-2282-420a-b1d6-1893fe4a74a6")
```

## Exploring a dataset
```{r}
colnames(nest_data)
glimpse(nest_data)
# head(nest_data)
unique(nest_data$species)
```

## Combining species dataset with predator_survey by species codes
```{r}
predator_names <- left_join(predator_survey, species, by = c("species" = "alpha_code"))
```

## Make it into a function
```{r}
assign_species_name <- function(df, species){
    return_df <- left_join(df, species, by = c("species" = "alpha_code"))
    return(return_df)
}
```

## adding more to the function to clean the dataset
```{r}
assign_species_name <- function(df, species){
    return_df <- left_join(df, species, by = c("species" = "alpha_code")) %>%
      select(year, site, date, common_name, count) %>% 
      filter(!is.na(common_name))
    
    return(return_df)
}
```

## Use function to clean all dataset
```{r}
predator_clean <- assign_species_name(predator_survey, species) 

nest_location_clean <- assign_species_name(nest_data, species)

eggs_clean <- assign_species_name(egg_measures, species)
```

## getting egg counts by year and species
```{r}
egg_counts <- eggs_clean %>% 
  group_by(year, site) %>% 
  count(name = "egg_number")
```

## getting predator counts by site
```{r}
predator_counts <- predator_clean %>% 
  select(-c(date, common_name)) %>% 
  group_by(year, site) %>% 
  count(name = "predator_number")

```

## merging predator and egg count by site
```{r}
counts <- left_join(egg_counts, predator_counts, by = c("site", "year"))
```


## plotting nest vs. predator counts
```{r}
ggplot(counts, aes(x = predator_number, y = egg_number)) + 
  geom_point()

ggplot(counts) +
  geom_col(aes(x = year, y = egg_number)) + 
  geom_point(aes(x = year, y = predator_number))
```

