---
title: "Spatial Data"
format: html
editor: visual
---

## Load libraries
```{r}
#| message: false

library(readr)
library(sf)
library(ggplot2)
library(leaflet)
library(scales)
library(ggspatial)
library(dplyr)
```

## Load shapefile from [KNB repository](https://dev.nceas.ucsb.edu/view/urn:uuid:6f07cb25-a4a1-48e8-95cb-74f532f3ce2d)
```{r}
# download URL
knb_url <- "https://dev.nceas.ucsb.edu/knb/d1/mn/v2/object/urn%3Auuid%3Aaceaecb2-1ce0-4d41-a839-d3607d32bb58"

# downloading zip file
download.file(url = knb_url, destfile = 'shapefile_demo_data.zip')

# unzipping
unzip('shapefile_demo_data.zip', exdir = 'data')

# removing zip file
file.remove('shapefile_demo_data.zip')
```

## Exploring the data using plot() and st_crs()
```{r}
# read in shapefile using read_sf()
ak_regions <- read_sf("data/ak_regions_simp.shp")
```

```{r}
# quick plot
plot(ak_regions)
```

```{r}
class(ak_regions)

head(ak_regions)

glimpse(ak_regions)

```

Look up current CRS with st_crs()
```{r}
st_crs(ak_regions)
# it is WGS 84
```

Changing CRS with st_transform() - changing to Alaska Albers projection, with an EPSG code of 3338.
```{r}
ak_regions_3338 <- ak_regions %>%
    st_transform(crs = 3338)

st_crs(ak_regions_3338)

plot(ak_regions_3338)
```

## sf & the Tidyverse
```{r}
# returns the names of all the columns in dataset
colnames(ak_regions_3338)

# can see how when selecting columns, the geometry is sticky and stays with the dataset
ak_regions_3338 %>%
    select(region)
```

```{r}
unique(ak_regions_3338$region)

ak_regions_3338 %>%
    filter(region == "Southeast")
```

## Exercise: How many people live in each of these Alaska regions?
Read in alaska population data
```{r}
#| message: false

# read in population data
pop <- read_csv("data/alaska_population.csv")
```

Turn population data into a spatial object
```{r}
pop_4326 <- st_as_sf(pop,
                     coords = c('lng', 'lat'),
                     crs = 4326,
                     remove = F)

head(pop_4326)
```

Transform to EPSG 3338
```{r}
pop_3338 <- st_transform(pop_4326, 
                         crs = 3338)
```

Left join with ak_regions_3338 using st_join() - for more info on different types of joins: ?st_within()
```{r}
pop_joined <- st_join(pop_3338, 
                      ak_regions_3338, 
                      join = st_within)

head(pop_joined)
```

Calculate total population using group_by() and summarize()
```{r}
pop_region <- pop_joined %>%
  as.data.frame() %>%  # will remove geometry, change from sf to dataframe
  group_by(region) %>%
  summarise(total_pop = sum(population))

head(pop_region)
```

Add total population info back to the shapefile
```{r}
pop_region_3338 <- left_join(ak_regions_3338, 
                             pop_region, 
                             by = "region")

# check
head(pop_region_3338)
plot(pop_region_3338["total_pop"])
```
*sf & Tidyverse note:*

The group_by() and summarize() functions can also be used on sf objects to summarize within a dataset and combine geometries. Many of the tidyverse functions have methods specific for sf objects, some of which have additional arguments that wouldn’t be relevant to the data.frame methods. You can run ?sf::tidyverse to get documentation on the tidyverse sf methods.

Calculate the population by Alaska management area
```{r}
pop_mgmt_3338 <- pop_region_3338 %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop))

plot(pop_mgmt_3338["total_pop"])
```
Notice that the region geometries were combined into a single polygon for each management area.
If we don't want them to combine, set do_union = F
```{r}
pop_mgmt_3338 <- pop_region_3338 %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop), do_union = F)

plot(pop_mgmt_3338["total_pop"])
```

Saving spatial object with write_sf()
```{r}
write_sf(pop_region_3338, "data/ak_regions_population.shp")
```

## Visualize with ggplot2
```{r}
# total population map, with gradient colors
ggplot(pop_region_3338) +
    geom_sf(aes(fill = total_pop)) +
    labs(fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) +
    theme_bw()
```
Adding in river data
```{r}
rivers_3338 <- read_sf("data/ak_rivers_simp.shp")
st_crs(rivers_3338)
```

```{r}
ggplot() +
    geom_sf(data = pop_region_3338, aes(fill = total_pop)) +
    geom_sf(data = rivers_3338, aes(linewidth = StrOrder)) +
    scale_linewidth(range = c(0.05, 0.5),  guide = "none") +
    labs(title = "Total Population by Alaska Region", fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) +
    theme_bw() 

```
## Incorporate basemaps with ggspatial
```{r}
ggplot(data = pop_3338) +
    ggspatial::annotation_map_tile(type = "osm", zoom = 4) + # higher zoom values are more detailed 
    geom_sf(aes(color = population),
            fill = NA) +
    scale_color_continuous(low = "darkkhaki",
                           high =  "firebrick",
                           labels = comma)
```
## Visualize sf objects with leaflet

Define a leaflet projection for Alaska Albers
```{r}
epsg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS",
    code = "EPSG:3338",
    proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
    resolutions = 2 ^ (16:7)
)
```

Leaflet requires an unprojected coordinate system, so use st_transform() to get back to WGS84
```{r}
pop_region_4326 <- pop_region_3338 %>% 
    st_transform(crs = 4326)
```

Plotting
```{r}
m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(data = pop_region_4326,
                fillColor = "gray",
                weight = 1)

m
```

Adding more details
```{r}
# creating color palette based on total population
pal <- colorNumeric(palette = "Reds", domain = pop_region_4326$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_region_4326,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1,
        label = ~ region
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_region_4326$total_pop),
        title = "Total Population"
    )

m
```

Adding communities with pop-ups
```{r}
# creating color palette based on total population
pal <- colorNumeric(palette = "Reds", domain = pop_region_4326$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_region_4326,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1
    ) %>%
    addCircleMarkers( # adding markers for cities
        data = pop_4326,
        lat = ~ lat,
        lng = ~ lng,
        radius = ~ log(population / 500), # arbitrary scaling
        fillColor = "gray",
        fillOpacity = 1,
        weight = 0.25,
        color = "black",
        label = ~ paste0(pop_4326$city, ", population ", comma(pop_4326$population))
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_region_4326$total_pop),
        title = "Total Population"
    )

m
```

