---
title: "Cleaning & Wrangling"
format: html
editor: visual
---

```{r}
#| message: false

library(dplyr)
library(tidyr)
library(readr)
```

# Data Sources

## Read in our data

```{r}

catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Explore

```{r}
#Prints the column names of my data frame
colnames(catch_original)

# First 6 lines of the data frame
head(catch_original)

# Summary of each column of data
summary(catch_original)

# Prints unique values in a column (in this case, the region)
unique(catch_original$Region)

#View() in console
```

# Clean and Reshape Data

refer to the book

%\>% Apple shortcut: Cmd + Shift + M Windows shortcut: Ctrl + Shift + M

## Using select() function

```{r}
catch_data <- catch_original %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
head(catch_data)

catch_data <- catch_original %>% 
  select(-All, -notesRegCode)
head(catch_data)
```

## check column types

```{r}
glimpse(catch_data)
# does anything seem amiss?
# we see that the values for 'Chinook' are actually character, and not a number
```

## Replace values in a column using mutate()

```{r}
# change chinook column to numeric
catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))
# we get an error though, can anyone figure out why?
```

```{r}
# find NA values
is.na(catch_clean$Chinook)

i <- which(is.na(catch_clean$Chinook))
i

catch_data[i,]

catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(condition = Chinook == "I",
                           true = "1",
                           false = Chinook),
         Chinook = as.numeric(Chinook))

# double-check
catch_clean[i,]

```

## Changing shape using pivot_longer() and pivot_wider()

```{r}
head(catch_clean)

catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
              values_to = "catch")
head(catch_long)

```

```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = species,
              values_from = catch)

head(catch_wide)
```

## Rename columns with rename()

```{r}
catch_long <- catch_long %>%
  rename(catch_thousands = catch)
head(catch_long)

```

```{r}
catch_long <- catch_long %>% 
  mutate(catch = catch_thousands * 1000) %>% 
  select(-catch_thousands)

head(catch_long)
```

## Summary stats using group_by() and summize()

```{r}
# get average catch by region
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch))
head(mean_region)

```

```{r}
# number of rows for each group
n_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(n = n())
head(n_region)

```

## Filtering rows using filter()

```{r}
sse_catch <- catch_long %>% 
  filter(Region == "SSE")
head(sse_catch)
```

### exercise

```{r}
# filter to catches over one million fish
catch_million <- catch_long %>% 
  filter(catch > 1000000)

# filter to just Chinook for SSE
chinook_sse <- catch_long %>% 
  filter(Region == "SSE" & species == "Chinook")

# note on ==, a logical test that if true keeps the row

head(chinook_sse)
```

## Sorting data using arrange()

```{r}
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(mean_catch)
head(mean_region)

mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))
head(mean_region)

```

## split and combine values in columns

```{r}
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAU-039"))
head(sites_df)

split_df <- sites_df %>% 
  separate(site, c("island", "site_number"), "-")
head(split_df)

```

### Exercise

```{r}
cities_df <- data.frame(city = c("Junea AK",
                                 "Sitka AK",
                                 "Anchorage AK"))

# split into two columns: city and state_code:
cities_df <- cities_df %>% 
  separate(city, c("city", "state_code"), " ")

```

```{r}
date_df <- data.frame(
  year = c("1930", "1930", "1930"),
  month = c("12", "12", "12"),
  day = c("14", "15", "16")
)
head(date_df)

dates_unite <- date_df %>% 
  unite(date, year, month, day, sep = "-")
head(dates_unite)

```

## Now all together:

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch * 1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)
```

## Save your results with write_csv()

```{r}
write_csv(mean_region, here::here("data/mean_catch_by_region.csv"))
```
